---
title: "Individual participant data meta-analysis. When? Why? How? A scoping review"
author: "Michail Belias"
date: "May 13, 2019"
header-includes:
    - \usepackage{booktabs}
    - \setlength\parindent{24pt}\setlength{\parskip}{0.0pt plus 1.0pt}
    - \usepackage{longtable}
    - \usepackage{array}
    - \usepackage{multirow}
    - \usepackage{wrapfig}
    - \usepackage{float}
    - \usepackage{colortbl}
    - \usepackage{pdflscape}
    - \usepackage{tabu}
    - \usepackage{threeparttable}
    - \usepackage{threeparttablex}
    - \usepackage[normalem]{ulem}
    - \usepackage{makecell}
    - \usepackage{xcolor}
output:
  pdf_document: default
  word_document:
    reference_docx: templatePaper.docx
  html_document: default
bibliography: Style and bibliography/bibliography2.bib
always_allow_html: yes
---

```{r loading the libraries and fixing the pander enviroment echo=F, warning=FALSE, message=FALSE}

library(knitr)
library(tinytex)
library(pander)
library(magick)
library(pander)
library(readxl)
library(dplyr)
library(ggpubr)
library(kableExtra)

options(tinytex.verbose = TRUE)
panderOptions('table.alignment.default', function(df)
    ifelse(sapply(df, is.numeric), 'right', 'left'))
panderOptions('table.split.table', Inf)
panderOptions('big.mark', ",")
panderOptions('keep.trailing.zeros', TRUE)
panderOptions("table.alignment.default", value = "left")

opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',echo=F, warning=FALSE, message=FALSE)


```


```{rloading the data-set and replacing 'NA' with NAs}

IPD_MA <- read_xlsx("IPD-MA Cochrane papers/6. Data/New search in PubMed.xlsx",  sheet = 1)
IPD_MA[IPD_MA == "NA"] <- NA
IPD_MA <- as.data.frame(IPD_MA)
```

\newpage

######

# Abstract
## Background
Individual participant data(IPD) meta-analysis(MA) is considered the gold standard for evidence based inference. It is well established that IPD-MA offers great advantages compared to aggregate MA and single studies, thus systematic reviews and methodological articles have been conducted in order to investigate current practice and propose guidance, on how to analyse and report an IPD-MA. Since the last systematic review in 2015 a new statistical approach to investigate effect modification in IPD-MA, new modelling approaches and a new Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) for IPD guideline have been published.

**Objective:** To conduct a scoping review of IPD-MA and summarise their reporting quality and statistical approaches used. Consequently, we aim to inform how IPD-MA are performed, what is their goal, which statistical approach they use and whether reporting was described according to PRISMA guidelines. 

## Methods 
We searched MEDLINE, PubMed and Cochrane IPD-MA Library for IPD-MAs related articles published the last five years. We screened the titles and abstracts and extracted where possible the size of the meta-analysis, their primary goal, type of outcome(s), study designs, statistical analysis and modelling approaches performed. Subsequently we excluded diagnostic, network, predictive and opportunistic IPD-MAs and articles with poor reporting and sampled the remaining for full text consideration. Finally, we assessed the quality of -within full text- reporting according to the PRISMA-IPD guidelines.

## Results
Our search resulted in 1538 articles, after exclusion criteria we ended with *702*. We sampled 100 and considered their full texts. IPD-MAs have seen considerable growth over the last five years. Random-effects are most often used and one-stage are almost equally performed as two-stage approaches. Most IPD-MAs have not clearly stated the goal, statistical approach and characteristics of their meta-analysis in their abstract and title. 

 
*Temporary*

**Most of the IPD-MAs had as a goal to investigate for subgroups effects.**
**Reporting type of **

## Conclusions
*Not yet*


\newpage
######

# Introduction

####### Meta-analysis (MA) is a statistical method that involves combining information from multiple sources. While initially meta-analyses were limited in aggregated data (AD) in the early 1990s individual participant data meta-analysis (IPD-MA or IPDMA) was introduced. In IPD-MA the participant level information is available and therefore evidence from multiple studies can be analysed as a whole. Collecting the IPD may be a difficult and time consuming task, but IPD-MA offers great opportunities [@van_Walraven_2010] that in AD-MA and single studies may not be possible. Therefore, IPD-MA is considered the gold standard in evidence synthesis [@Stewart_1993 ; @Stewart_1995 ;@Stewart_2002]. 

#######   Due to these advantages literature is present to provide guidance:1) over statistical approaches that should be applied and 2) how to report an IPD-MA. Statistical approaches vary depending on the goal of an IPD-MA. For instance, Simmonds et al.[@Simmonds_2015] showed that IPD-MAs are frequently performed in order to detect treatment effect modification and identified four approaches were mostly used. Specifically, an AD-MA approach called meta-regression and three IPD-MA approaches, per-subgroup meta-analysis, meta-analysis of interaction terms and one-stage IPD-MA were typically applied. Combinations of these approaches have been compared to provide guidance on which to choose. Simmonds and Higgins [@Simmonds_2007] mathematically proved that, one-stage IPD-MA is always more or equally powerful than meta-analysis of interaction terms and meta-regression. Nevertheless, in order to end up with a closed mathematical form for power they made unrealistic assumptions. Furthermore, Fisher et al. [@Fisher_2011] critically reviewed all four approaches and concluded that one-stage IPD-MA allows for more complex analysis, but is more difficult to perform than pooling within-trial interaction terms. Finally, Hua et al. [@Hua_2016] advocated that one-stage IPD-MAs using mixed-effects modelling should also centre the effect modifiers to their mean, in order to separate across and within trial information and therefore account for ecological bias. 

#######   IPD-MA may be also performed in order to investigate risk-factors. In that case typically observational studies are included and statistical analysis involves multi-variable modelling and investigating non-linear functional forms. For risk factor assessment Sauerbrei and Royston [@Sauerbrei_2011] suggested the use of a two stage approach involving polynomials (fractional polynomial or FPs) with degrees driven from a set of powers. In the first stage a FP is selected and at a second stage their estimates are pooled using weighted meta-analysis. In a subsequent article they extended their approach to also include interactions [@royston_interaction_2013]. Finally, White et al [@White_2018] compared two pooling methods to be applied in the second stage of FPs 1) 'metacurve' a point-wise regression line averaging technique and 2) 'mvmeta' a multi-variate meta-analysis coefficient pooling technique. Nevertheless, they mentioned that both techniques may be applied in other types of fitting techniques such as regular polynomials and splines.

#######    On the other hand, an extended version of Preferred Reporting Items for Systematic Review and Meta-Analyses (PRISMA) for individual participant data (IPD) [@Stewart_2015] has been developed. PRISMA-IPD offers guidance in the reporting of the title and abstact, and the full article (introduction, methods, results, discussion and funding). For instance, researchers should identify in the title their report as a systematic review and/or meta-analysis of individual participant data. Furthermore, they should provide in the abstract a clear background statement, describe the eligibility criteria and search strategy, provide the number of studies and participants, report the summary effect estimates and measures of heterogeneity and finally, they should state the main strengths and limitations, interpret the results and report funding sources. Subsequently, reporting should be more detailed in the full text. The differences compared to the original PRISMA [@Moher_2009] lie in the : 1) methods of checking the integrity of the IPD (such as pattern of randomization, data consistency, baseline imbalance, and missing data), (2) reporting any important issues that emerge, and (3) exploring variation (such as whether certain types of individual benefit more from the intervention than others). 


#######   Some of the aforementioned properties of an IPD-MA have been investigated in systematic reviews over the years. For instance, Simmonds et al [@Simmonds_2005] identified 44 IPD-MAs performed during 2000-2005 time period and summarized 1) whether IPD-MAs obtained all the data they sought 2) if they clearly described the statistical approaches they used 3) if the effects of co-variables have also been investigated and 4) their medical field. On a subsequent paper, 10 years later Simmonds et al. [@Simmonds_2015] identified 1371 potential IPD-MAs published during 2010-2015 time period, sampled 184 of them and included 100 IPD-MAs that provided full texts. Then along with the topics investigated in the initial paper they investigated also the quality of IPD-MA reporting. Riley et al. [@Riley_2010] identified 383 IPD-MAs performed from instance until 2009 and summarised only: 1) their medical field topic and 2) whether they assessed risk or prognostic factors. Finally, Schuit and Ioannidis [@Schuit_2018] identified 327 IPD-MAs performed from inception until 2014. They restricted their interest in treatment effects investigation and showed that 


####### As far as we know no Scoping or systematic review has evaluated the complience to the new statistical approaches or the new PRISMA guidelines described above. Therefore, our goal is to conduct a scoping review of IPD-MA from 2015 and onwards and summarise their aforementioned properties. Consequently, we aim to inform why and how IPD-MA are performed, which statistical approaches are preferred and whether they are clearly described according to the PRISMA-IPD guidelines. 

\newpage
######

# Methods 

####### This study is a scoping review of current practices in IPD-MAs. We report our study according to the Preferred Reporting Items for Systematic reviews and Meta-Analyses extension for Scoping Reviews (PRISMA-ScR) guidelines. *(bold are checkpoints in the PRISMA-ScR check-list)*

**Protocol and registration**
No formal protocol exists for this study. 

**Information sources**
A MEDLINE, PubMed and Cochrane library search was performed in $1^{st}$ of May using the following search terms: _(("Meta-Analysis"[Publication Type] OR meta-analys$*$[tiab] OR metaanalys$*$[tiab]) AND (individual participant[tiab] OR individual participants[tiab] OR individual participant's[tiab] OR individual patient[tiab] OR individual patients[tiab] OR individual patient's[tiab] OR individualized participant[tiab] OR individualized participants[tiab] OR individualized participant's[tiab] OR individualized patient[tiab] OR individualized patients[tiab] OR individualized patient's[tiab] OR individualised participant[tiab] OR individualised participants[tiab] OR individualised participant's[tiab] OR individualised patient[tiab] OR individualised patients[tiab] OR individualised patient's[tiab]) AND data[tiab]) OR IPDMA[tiab] OR IPD-MA[tiab] AND ("2015/01/01"[PDat] : "2019/05/01"[PDat])_ 

**Eligibility criteria**
We included studies describing an IPD-MA published between 01/01/2015 and 01/05/2019. We removed duplicate papers and from series of articles we included only the most recent. Our scoping review has been performed in two stages, one evaluating the title and abstract of all IPD-MAs and one evaluating the full text of a subset of them. Specifically, in the first stage we considered all available IPD-MAs and screened their titles and abstracts. Thereto, we extracted information -if present- over the year, medical field, number of included studies and participants, the goal, types of outcomes and statistical approaches preferred. Subsequently, we excluded diagnostic test and predictive IPD-MAs. Other reasons for further exclusion were: 1) full text was not published in English, 2) IPD-MAs were given as examples (for instance in methodological, health-technology assessments and cost-effectiveness studies) 3) we had no access to the full text and 4) the studies were protocols. Finally, we sampled 100 of these eligible studies and extracted information over the statistical analysis and evaluated the reporting quality according to the PRISMA-IPD guidelines.

**Data items**

We summarised our results using descriptive statistics such as frequency tables and bar-plots.


\newpage

# Results

Our search identified 1538 potential records. We removed 61 duplicates and series of articles. We screened the titles and abstracts of the remaining 1477 studies. Approximately 36% of the papers were not identified as a systematic review, or meta-analysis or meta-analysis of individual participant data.



```{r}
IPD_MA%>% 
  filter(`General Medical Field` != 'Statistical')%>%
  group_by(`General Medical Field`,`Type of study in title`)%>%
  summarise(Frequency =  n())%>%
  arrange(`General Medical Field`)%>%
  mutate(Percentage = paste(round(Frequency*100/sum(Frequency),2), "%", sep = "") ) %>%
  pander(style = "grid",keep.line.breaks = TRUE, caption = "Table 1. Percentage of articles indicating their type of study in the title") %>%
  select_(`General Medical Field`,`Type of study in title`,Percentage)
```

We show an increasing trend over the years of articles involved with IPD-MA. IPD methods seem to gain popularity in medical fields such as pregnancy and childbirth, mental health and neurology 

```{r}

IPD_Year_Frequency = IPD_MA %>%
  group_by(Year)%>%
  summarise(Frequency=n())

IPD_Year_Frequency$estimated = rep("No")
IPD_Year_Frequency[6,] = c(2019, 350, "Yes" )
IPD_Year_Frequency$estimated =  factor(IPD_Year_Frequency$estimated, levels = c("Yes","No"))

ggbarplot(data = IPD_Year_Frequency, x =  "Year", y = "Frequency", fill = "estimated",label = F,palette = "jco",
          ggtheme = theme_minimal(), title = "Number of IPD-MAs performed per year", caption =  "Figure 1.",xlab = "Year" ) + 
    theme(plot.title = element_text(hjust = 0.5,size = 18, face = "bold.italic") ,
          plot.subtitle = element_text(hjust = 0.5,size = 12, face = "bold.italic"))+ geom_text( aes(label=Frequency), vjust=c(1.5,1.5,1.5,1.5,1.5,-13))

```



*Title results*


*Synthesis of results*





```{r}
IPD_MA%>% 
  filter(`General Medical Field` !=  "Statistical") %>%
  filter(!is.na(`General Medical Field`))%>%
  group_by(`General Medical Field`)%>%
  summarise(Frequency =  n())%>%
  arrange(desc(Frequency))%>%
  mutate(Percentage = paste(round(Frequency*100/sum(Frequency),2), "%", sep = "") )%>%
  head(n= 5)%>%
pander(style = "grid",keep.line.breaks = TRUE, caption = "Table 2. Individual participant meta-analysis per medical field (top 5 reported)" )
```
####### The medical fields with the most articles were Cardiovascular diseases, cancer, neurology and mental health.


  
\newpage 
#####



```{r}

names(IPD_MA) = gsub(x = names(IPD_MA), pattern = " ", replacement = ".")
IPD_MA$Number.of.Studies= as.numeric(IPD_MA$Number.of.Studies)

p = gghistogram(data = IPD_MA, x =  "Number.of.Studies", y = "..count..",bins = 30, palette = "jco",
            ggtheme = theme_minimal(), title = "Number of trials included per meta-analysis", add = "median", xlab = "Trials per Meta-analysis" )  + annotate(geom="text", x=100, y=20, label= paste("median:", median(IPD_MA$Number.of.Studies,na.rm = T)), color="red") + theme(plot.title = element_text(hjust = 0.5,size = 18, face = "bold.italic")) 

p
```


```{r}


IPD_MA$Number.of..Patients = as.numeric(IPD_MA$Number.of..Patients)

gghistogram(IPD_MA, x =  "Number.of..Patients", y = "..count..",bins = 50, palette = "jco", 
            ggtheme = theme_minimal(), title = "Number of participants included per meta-analysis", add = "median", xlab = "Participants per Meta-analysis" )  + annotate(geom="text", x=5000, y=20, label= paste("median:", median(IPD_MA$Number.of..Patients,na.rm = T)), color="red") + theme(plot.title = element_text(hjust = 0.5,size = 18, face = "bold.italic")) +  xlim(0,25000)

```



```{r}


IPD_MA$Number.of..Patients = as.numeric(IPD_MA$Number.of..Patients)

IPD_Outcome_Frequency =  unlist(strsplit(gsub("[\r\n]" ,"",IPD_MA$Type.of.primary.outcome),";"))
IPD_Outcome_Frequency =  gsub(" ", "",IPD_Outcome_Frequency )
IPD_Outcome_Frequency =  table(IPD_Outcome_Frequency)


pander( IPD_Outcome_Frequency[names(IPD_Outcome_Frequency) %in% c("Binary", "Continuous", "Time-to-event")], 
        caption = "Table 3. Type of outcomes investigated",style= "grid")
```

\newpage 
#####

```{r}

Het_table <- read_excel("IPD-MA Cochrane papers/6. Data/Heterogeneity.xlsx")
colnames(Het_table)[1] = "  "

Het_table[is.na(Het_table) ] = " "

pander(Het_table, caption = "Table 4. Statistical approaches performed to detect treatment-effect modification", style= "grid")
```



\newpage

######

# Discussion


\newpage

######
# References




